---
title: "ERA LTE Data Preparation with Climate Merge"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
---

> Builds the ERA long-term-experiment (LTE) dataset: the latest public **ERA compiled** release, restricted to a curated list of LTE publications, joined to ERA's seasonal climate statistics. Everything is downloaded from public URLs (no credentials, no R packages beyond the tidyverse + arrow). Outputs are written to `data/`; the column dictionary is `data/data_dictionary.csv` and the build provenance is `data/VERSION.json`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

```{r packages}
library(dplyr); library(readr); library(stringr); library(purrr); library(arrow); library(jsonlite)
dl <- "downloaded_data"; dir.create(dl, showWarnings = FALSE)
s3 <- "https://digital-atlas.s3.amazonaws.com/era"
fetch <- function(url, name = basename(url)) {                # download once, then reuse the local copy
  f <- file.path(dl, name); if (!file.exists(f)) download.file(url, f, mode = "wb", quiet = TRUE); f }
```

## 1. ERA compiled data (latest public release)

The ERA team publishes versioned releases under `era/data/releases/`; `latest.json` names the current one. One row = one control-vs-treatment comparison for one outcome.

```{r era}
release  <- fromJSON(paste0(s3, "/data/releases/latest.json"))$latest
rel_url  <- paste0(s3, "/data/releases/version=", release, "/")
manifest <- fromJSON(paste0(rel_url, "manifest.json"))
era <- read_parquet(fetch(paste0(rel_url, "era_compiled.parquet"), paste0("era_compiled_", release, ".parquet")))
cat("ERA release", release, ":", nrow(era), "rows,", n_distinct(era$Code), "publications\n")
```

## 2. Curated LTE list

The registry lives in the ERA extraction repo (`CIAT/ERA_dev`). It is semicolon-delimited, Latin-1 encoded, has trailing empty columns and a few broken rows, so it is cleaned here and the cleaned copy is shipped as `data/unique_ltes_clean.csv`. Rows that cannot be used are listed in `data/unique_ltes_unresolved.csv` instead of being dropped silently.

```{r registry}
reg_url <- "https://github.com/CIAT/ERA_dev/raw/refs/heads/main/data_entry/long_term_experiments/unique.ltes.csv"
reg_raw <- read_delim(fetch(reg_url), delim = ";", locale = locale(encoding = "latin1"),
                      name_repair = "minimal", show_col_types = FALSE)
reg0 <- reg_raw[, nzchar(names(reg_raw))] |> rename_with(str_trim) |> mutate(across(where(is.character), str_trim))
ok   <- str_detect(coalesce(reg0$Code, ""), "^[A-Z]{2}[0-9]{4}([.][0-9]+)?$")   # a real ERA publication code
reg  <- reg0[ok, ] |> distinct(LTE.ID, Code, Site.ID, .keep_all = TRUE)
unresolved <- bind_rows(`malformed row` = reg0[!ok, ], `code not in ERA compiled` = filter(reg, !Code %in% era$Code), .id = "issue")
write_csv(reg, "data/unique_ltes_clean.csv"); write_csv(unresolved, "data/unique_ltes_unresolved.csv")
cat(nrow(reg), "registry rows:", n_distinct(reg$LTE.ID), "LTEs,", n_distinct(reg$Code), "publications;", nrow(unresolved), "unresolved rows\n")
```

## 3. Filter to LTE publications and attach the LTE identifier

One LTE can have several publications and one publication can report several LTEs (multi-site papers). `LTE.ID` is attached by publication **and site** where the site names agree, otherwise by publication when that is unambiguous. Rows that remain without `LTE.ID` are counted in the build report.

```{r filter}
lte_map <- distinct(reg, Code, Site.ID, LTE.ID)
by_site <- distinct(lte_map, Code, Site.ID, .keep_all = TRUE)
by_code <- lte_map |> group_by(Code) |> filter(n_distinct(LTE.ID) == 1) |> distinct(Code, LTE.ID)
lte <- era |> filter(Code %in% reg$Code) |>
  left_join(by_site, by = c("Code", "Site.ID")) |>
  left_join(by_code, by = "Code", suffix = c("", "_code")) |>
  mutate(LTE.ID = coalesce(LTE.ID, LTE.ID_code), .keep = "unused")
cat(nrow(lte), "LTE rows,", n_distinct(lte$Code), "publications,", n_distinct(lte$LTE.ID), "LTEs;", sum(is.na(lte$LTE.ID)), "rows without LTE.ID\n")
```

## 4. Analysis-ready columns

-   **Yield in t/ha** (`yield_t_ha` for the treatment, `yield_c_t_ha` for the control) with an explicit `yield_basis`, so users never write their own unit map. Only crop-yield rows are converted.
-   **Aggregated years**: `M.Year` such as `1993..2003` means the value is a mean over several years or seasons; it is flagged and its first/last year parsed.
-   **Arm ids**: `Index` identifies the *pair*. The same treatment arm appears once per control it is paired with, so `obs_id` (treatment arm × site × year × crop × experimental unit × soil depth) and `ctrl_id` identify the arms themselves. Use them to de-duplicate before any statistic.
-   **Date provenance**: whether the planting and harvest dates come from the publication.

```{r derive}
units <- tribble(~Units, ~yield_factor, ~yield_basis,
  "Mg/ha", 1, "as reported", "kg/ha", 0.001, "as reported", "q/ha", 0.1, "as reported",
  "Mg DM/ha", 1, "dry matter", "kg DM/ha", 0.001, "dry matter",
  "Mg/ha/yr", 1, "per year", "kg/ha/yr", 0.001, "per year")
lte <- lte |> left_join(units, by = "Units") |>
  mutate(is_yield     = Out.SubInd == "Crop Yield",
         yield_t_ha   = if_else(is_yield, MeanT * yield_factor, NA),
         yield_c_t_ha = if_else(is_yield, MeanC * yield_factor, NA),
         yield_basis  = if_else(is_yield, yield_basis, NA),
         m_year_is_aggregate = str_detect(M.Year, fixed("..")),
         m_year_first = as.integer(str_extract(M.Year, "[0-9]{4}")),
         m_year_last  = as.integer(str_extract(M.Year, "[0-9]{4}(?!.*[0-9]{4})")),
         obs_id  = paste(Site.Key, M.Year, Product.Simple, EU, TID, Upper, Lower, sep = "|"),
         ctrl_id = paste(Site.Key, M.Year, Product.Simple, EU, CID, Upper, Lower, sep = "|"),
         is_livestock = Product.Type %in% "Animal" | Version %in% "skinny_cow_2022",
         Duration = if_else(Duration < 0, NA, Duration),                 # negative durations are data-entry errors
         planting_date_source = if_else(is.na(Plant.Start), "none", "reported"),
         harvest_date_source  = if_else(is.na(Harvest.End), "none", "reported"),
         era_release = release, era_provenance = manifest$datasets$pipeline_provenance[1]) |>
  select(-Lat, -Lon, -yield_factor, -is_yield)                          # Lat/Lon are empty legacy columns
```

## 5. Seasonal climate

ERA's geodata pipeline computes growing-season climate statistics per site × year × crop × planting window (`clim_stats_*.RData`). Two sets are used, in order:

1.  **PDate** — windows from planting dates reported in the publication (`PDate.SLen.Data`), matched on site, year, crop and planting window.
2.  **EcoCrop** — generic crop-calendar windows (`PDate.SLen.EcoCrop`), used only where step 1 finds nothing, matched on site, year and crop. Treat these rows with more caution: the window is not the study's own.

`climate_date_source` records which set matched; `climate_window_start/end` is the window the statistics were actually computed over (for EcoCrop rows it can differ from a reported `Plant.Start`). Climate values are rounded to 3 decimals.

```{r climate}
listing  <- paste(readLines("https://digital-atlas.s3.amazonaws.com/?list-type=2&prefix=era/geodata/clim_stats", warn = FALSE), collapse = "")
clim_key <- str_extract_all(listing, "era/geodata/clim_stats_[0-9.-]+[.]RData")[[1]] |> sort() |> last()
cs <- (\(f) { e <- new.env(); load(f, envir = e); get(ls(e)[1], e) })(fetch(paste0("https://digital-atlas.s3.amazonaws.com/", clim_key)))

prep <- function(tabs, keys) {                                            # one wide table per window set
  tabs <- set_names(tabs, recode(names(tabs), rainfall = "rain", temperature = "temp"))
  imap(tabs, \(x, nm) as_tibble(x) |> mutate(prod_key = str_to_lower(str_trim(Product))) |>
         select(-any_of(c("row_index", "window", "EU", "Product"))) |>
         distinct(across(all_of(keys)), .keep_all = TRUE) |> rename_with(\(v) paste0(nm, "_", v), -all_of(keys))) |>
    reduce(full_join, by = keys) }
keys_p <- c("Site.Key", "M.Year", "prod_key", "Plant.Start", "Plant.End"); keys_e <- keys_p[1:3]
clim_p <- prep(cs$PDate.SLen.Data, keys_p); clim_e <- prep(cs$PDate.SLen.EcoCrop, keys_e)
clim_cols <- setdiff(names(clim_p), keys_p)

lte <- lte |> mutate(prod_key = str_to_lower(str_trim(Product.Simple))) |> left_join(clim_p, by = keys_p) |>
  mutate(climate_date_source = if_else(is.na(gdd_gdd_opt), NA, "PDate"),
         climate_window_start = if_else(is.na(gdd_gdd_opt), as.Date(NA), Plant.Start),
         climate_window_end   = if_else(is.na(gdd_gdd_opt), as.Date(NA), Plant.End))
fill <- lte |> filter(is.na(climate_date_source)) |> select(-all_of(clim_cols)) |> left_join(clim_e, by = keys_e) |>
  mutate(climate_date_source  = if_else(is.na(gdd_gdd_opt), NA, "EcoCrop"),
         climate_window_start = gdd_Plant.Start, climate_window_end = gdd_Plant.End,
         planting_date_source = if_else(planting_date_source == "none" & !is.na(gdd_Plant.Start), "EcoCrop", planting_date_source),
         Plant.Start = coalesce(Plant.Start, gdd_Plant.Start), Plant.End = coalesce(Plant.End, gdd_Plant.End)) |>
  select(-ends_with("_Plant.Start"), -ends_with("_Plant.End"))
lte <- bind_rows(filter(lte, !is.na(climate_date_source)), fill) |>
  mutate(harvest_date_source = if_else(harvest_date_source == "none" & !is.na(rain_Harvest.End), "ERA climate window", harvest_date_source),
         across(all_of(clim_cols) & where(is.numeric), ~ round(.x, 3))) |>
  select(-prod_key)
print(count(lte, climate_date_source, planting_date_source))
```

Rows without climate are exported with the reason, so they can be worked on rather than guessed at.

```{r unmatched}
unmatched <- lte |> filter(is.na(climate_date_source)) |>
  mutate(reason = case_when(is.na(Latitude) ~ "no coordinates", is.na(Product.Simple) ~ "no crop",
                            m_year_is_aggregate ~ "aggregated M.Year", is.na(Plant.Start) ~ "no planting date",
                            TRUE ~ "no climate row for this key")) |>
  count(Site.Key, M.Year, Product.Simple, Plant.Start, Plant.End, reason, name = "n_rows")
write_csv(unmatched, "data/unmatched_climate_keys.csv"); print(count(unmatched, reason, wt = n_rows))
```

## 6. One row per arm measurement (`lte_arms.csv`)

`lte_final.csv` keeps ERA's pair structure (needed for effect sizes). For time series, yield-vs-climate plots or any per-treatment statistic, use `lte_arms.csv`: every control and treatment arm once per outcome, year and distinct measurement, with the climate attached. `arm_role` is `T`, `C` or `C/T` (an arm that is the control in some pairs and the treatment in others); `n_pairs` is the number of pairs the identical measurement came from; `n_means` \> 1 flags an arm with several distinct values for the same outcome (different sampling dates or data locations in the paper: see `DataLoc`, `Duration`, `Plant.Start`).

```{r arms}
common <- c("LTE.ID", "Code", "Site.Key", "Site.ID", "Country", "Latitude", "Longitude", "M.Year", "m_year_first", "m_year_last",
            "m_year_is_aggregate", "Product.Simple", "EU", "Out.SubInd", "Outcode", "Units", "yield_basis", "Mean.Error.Type",
            "Upper", "Lower", "DataLoc", "Analysis.Function", "Duration", "Variety", "Rep", "Plot.Size",
            "Plant.Start", "Plant.End", "Harvest.Start", "Harvest.End", "planting_date_source", "harvest_date_source",
            "climate_date_source", "climate_window_start", "climate_window_end", "is_livestock", clim_cols)
arms <- bind_rows(
  T = lte |> transmute(across(all_of(common)), arm_id = obs_id,  arm_code = TID, arm_label = T.Descrip, mean = MeanT, mean_error = MeanT.Error, yield_t_ha, pair_index = Index),
  C = lte |> transmute(across(all_of(common)), arm_id = ctrl_id, arm_code = CID, arm_label = C.Descrip, mean = MeanC, mean_error = MeanC.Error, yield_t_ha = yield_c_t_ha, pair_index = Index),
  .id = "arm_role") |>
  group_by(arm_id, Outcode) |> mutate(n_means = n_distinct(mean)) |>
  group_by(arm_id, Outcode, mean) |> mutate(arm_role = paste(sort(unique(arm_role)), collapse = "/"), n_pairs = n()) |>
  slice(1) |> ungroup() |> relocate(arm_role, arm_id, n_pairs, n_means, pair_index, .before = 1)
write_csv(arms, "data/lte_arms.csv", na = "")
cat(nrow(arms), "arm measurements from", nrow(lte), "pairs;", sum(arms$n_means > 1), "rows belong to arms with more than one value for the same outcome\n")
```

## 7. Write the dataset, the build report and the provenance

```{r write}
lte <- lte |> relocate(LTE.ID, Code, Site.Key, Site.ID, Index, obs_id, ctrl_id, M.Year, Product.Simple, Out.SubInd, .before = 1)
write_csv(lte, "data/lte_final.csv", na = "")

report <- tibble::enframe(c(
  `ERA compiled rows`               = nrow(era),
  `LTE rows`                        = nrow(lte),
  `LTE publications`                = n_distinct(lte$Code),
  `LTEs (LTE.ID)`                   = n_distinct(lte$LTE.ID),
  `rows without LTE.ID`             = sum(is.na(lte$LTE.ID)),
  `sites (Site.Key)`                = n_distinct(lte$Site.Key),
  `countries`                       = n_distinct(lte$Country),
  `crop-yield rows`                 = sum(lte$Out.SubInd == "Crop Yield"),
  `crop-yield rows with yield_t_ha` = sum(!is.na(lte$yield_t_ha)),
  `rows with aggregated M.Year`     = sum(lte$m_year_is_aggregate),
  `planting date reported`          = sum(lte$planting_date_source == "reported"),
  `climate matched: PDate`          = sum(lte$climate_date_source %in% "PDate"),
  `climate matched: EcoCrop`        = sum(lte$climate_date_source %in% "EcoCrop"),
  `climate missing`                 = sum(is.na(lte$climate_date_source)),
  `livestock rows (flagged)`        = sum(lte$is_livestock),
  `arm measurements (lte_arms)`     = nrow(arms),
  `crop-yield arm measurements`     = sum(arms$Out.SubInd == "Crop Yield")), name = "metric", value = "value")
write_csv(report, "data/build_report.csv"); print(as.data.frame(report))

write_json(list(lte_release = "2026.1", built = format(Sys.Date()), era_release = release,
                era_provenance = manifest$datasets$pipeline_provenance[1], clim_stats = basename(clim_key),
                registry = reg_url, rows = nrow(lte), publications = n_distinct(lte$Code), ltes = n_distinct(lte$LTE.ID)),
           "data/VERSION.json", auto_unbox = TRUE, pretty = TRUE)
```

## 8. Column dictionary

ERA's own field descriptions (`data/era_compiled_fields.csv`, exported from the `eragri` package) plus the columns added here and the climate variables.

```{r dictionary}
derived <- tribble(~Field.Name, ~Description,
  "LTE.ID", "Long-term-experiment identifier from the curated registry (one LTE can span several publications).",
  "obs_id", "Treatment-arm id: Site.Key | M.Year | Product.Simple | EU | TID | Upper | Lower. Identical across the pairs an arm appears in.",
  "ctrl_id", "Control-arm id: Site.Key | M.Year | Product.Simple | EU | CID | Upper | Lower.",
  "yield_t_ha", "MeanT converted to t/ha for crop-yield rows (see yield_basis); NA for other outcomes or unconvertible units.",
  "yield_c_t_ha", "MeanC converted to t/ha for crop-yield rows.",
  "yield_basis", "'as reported' (fresh / standard moisture as published), 'dry matter', or 'per year' (annualised).",
  "m_year_is_aggregate", "TRUE when M.Year covers several years or seasons ('..' separator): the value is a multi-year mean.",
  "m_year_first", "First calendar year in M.Year.", "m_year_last", "Last calendar year in M.Year.",
  "is_livestock", "TRUE for animal-product rows (kept for completeness; not agronomy LTE data).",
  "planting_date_source", "'reported' (from the publication), 'EcoCrop' (generic crop calendar, filled here) or 'none'.",
  "harvest_date_source", "'reported', 'ERA climate window' (harvest end taken from the climate window) or 'none'.",
  "climate_date_source", "Which climate set matched: 'PDate' (reported planting window) or 'EcoCrop' (generic window); NA = no climate.",
  "climate_window_start", "Start of the window the climate statistics were computed over.",
  "climate_window_end", "End of that window.",
  "era_release", "ERA compiled release the row comes from.", "era_provenance", "ERA pipeline provenance string of that release.",
  "arm_role", "lte_arms only: T, C, or C/T when the arm is control in some pairs and treatment in others.",
  "arm_id", "lte_arms only: obs_id or ctrl_id of the arm.", "arm_code", "lte_arms only: ERA treatment code (TID/CID).",
  "arm_label", "lte_arms only: treatment description (T.Descrip/C.Descrip).", "mean", "lte_arms only: reported mean of the arm.",
  "mean_error", "lte_arms only: reported dispersion of the mean.",
  "n_pairs", "lte_arms only: number of pairs in lte_final.csv that carry this identical arm measurement.",
  "n_means", "lte_arms only: number of distinct values this arm has for this outcome and year (1 = a single measurement).",
  "pair_index", "lte_arms only: Index of the first pair the measurement appears in.") |> mutate(Source = "this repository")
clim_dict <- tibble(Field.Name = clim_cols) |> mutate(
  Description = case_when(
    str_ends(Field.Name, "Harvest.Start|Harvest.End") ~ "Harvest date bounding the climate window (per climate table).",
    str_starts(Field.Name, "gdd_")    ~ "Growing degree days in the window: suboptimal / optimal / above-optimal / above-maximum temperature ranges (crop-specific thresholds).",
    str_starts(Field.Name, "rain_")   ~ "Rainfall in the window: total (mm), reference ET, water balance, and dry-spell counts for thresholds of 0.1, 1 and 5 mm/day (days, share of days, longest run, runs of 5/10/15+ days).",
    str_starts(Field.Name, "temp_")   ~ "Temperature in the window: min/max/mean statistics and heat-stress days above 35, 37.5 and 40 C (days, share, longest run, runs of 5/10/15+ days).",
    str_starts(Field.Name, "eratio_") ~ "Evapotranspiration ratio (actual/potential) in the window: mean, median, min and drought-stress days below 0.5, 0.25 and 0.1.",
    str_starts(Field.Name, "logging_") ~ "Waterlogging indicators in the window (ERA geodata).",
    TRUE ~ "Climate statistic from ERA geodata."),
  Source = paste0("ERA geodata ", basename(clim_key)))
dict <- bind_rows(read_csv("data/era_compiled_fields.csv", show_col_types = FALSE), derived, clim_dict) |>
  filter(Field.Name %in% union(names(lte), names(arms))) |> distinct(Field.Name, .keep_all = TRUE)
write_csv(dict, "data/data_dictionary.csv")
cat(nrow(dict), "columns documented of", length(union(names(lte), names(arms))), "\n")
```
